Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-12-11 scagnostics.Rmd
Replacing melt by gather everywhere.

```{r load}
library(tourr)
library(tidyverse)
library(reshape2)
library(scagnostics)
library(gridExtra)
library(tictoc) #timer
library(mbgraphic) #Katrins package
library(GGally)
library(geozoo)
library(minerva) #MINE indices
library(kableExtra)
library(forcats)
```

@@

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine (and Spiral) dataset the parameter $x_6$ (and $x_5$) follows a very distinct distribution.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- melt(ds)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between a nuisance and a structured projection (solid lines: $n=1000$; dashed: $n=100$). Most of the indexes have peaks at the end of the sequence, indicating that they see the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection."}
#fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
fullResMeltPlanned <- fullResPlanned %>%
  mutate(convex = 1-convex) %>% 
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

```{r pipeFirstRun, fig.width=7, fig.height=7, out.width = "0.8\\textwidth", fig.cap="Scouting for the pipe using optimization method 2 on the TIC index. Other PPI values also shown for interest. Only the values for the target plane are shown because the optimization is done off-line. The TIC maximum value for this data is small,  even though it identifies the pipe. The holes, convex, skinny, and to some extent MIC, mirror the performance of TIC for this structure."}
if(!file.exists("cache/pipeFirstRun.rda")){
  pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
  pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
  save(pipeTourRes, pipeTourResMelt, file = "cache/pipeFirstRun.rda")
} else {
  load("cache/pipeFirstRun.rda")
}
newBasisPipeTour <- pipeTourRes %>%
  mutate(convex = 1-convex) %>% 
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  melt(id=c("t", "name", "size")) %>%
  filter(t %in% which(attributes(interpolate(pipeTour))$new_basis))
ggplot(newBasisPipeTour, aes(x=t, y=value)) +
  geom_point(mapping=aes(color=(variable=="TIC"))) + 
  geom_line(mapping=aes(color=(variable=="TIC"))) + 
  scale_colour_manual(values = c("black", "red")) +
  facet_wrap(~variable, ncol=1, strip.position = "right") +
  guides(color=FALSE) +
  ylim(c(0,1)) +
  xlab("Number of projections") +
  ylab("PPI value") 
```

```{r refinepipe, results="hide"}
if(!file.exists("cache/refinepipe.rda")){
  iLast <- length(pipeTourFull)
  fProj <- pipeTourFull[[iLast]]
  pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
  pipeTourFull2 <- as.list(interpolate(pipeTour2))
  pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
  pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
  save(pipeTour2, pipeTourFull2, pipeTourRes2, pipeTourResMelt2, file = "cache/refinepipe.rda")
} else {
  load("cache/refinepipe.rda")
}
```

@@

```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising the splines2d index for the Sine 1000 dataset. Anchor planes are marked by the blue vertical lines.", results="hide"}
set.seed(2018)
if(!file.exists("cache/findsine.rda")){
  sineResc <- as.matrix(sin1000)
  sineTour <- save_history(sineResc,
                          guided_tour(splineIndex()))
  sineTourFull <- as.list(interpolate(sineTour))
  sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
  sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
  save(sineResc, sineTour, sineTourFull, sineTourRes, sineTourResMelt, file = "cache/findsine.rda")
} else {
  load("cache/findsine.rda")
}
anchorIdx <- as.tibble(which(attributes(interpolate(sineTour))$new_basis))
sineTourResMelt <-sineTourRes %>%
  mutate(convex = 1-convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  melt(id=c("t", "name", "size"))
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line(mapping=aes(color=(variable=="splines2d"))) + 
  facet_wrap(~variable, ncol=1, scales = "free_y") +
  scale_colour_manual(values = c("black", "red")) +
  facet_wrap(~variable, ncol=1) +
  geom_vline(data=anchorIdx, mapping=aes(xintercept=value), color="blue", alpha=0.3) +
  guides(color=FALSE) +
  ylim(c(0,1))
```

@@

  return(sc)
}
if(!file.exists("cache/spiralScagBinning.rda")){
  scagBinningDf <- getScagComp(spiral1000, t2full)
  scagMelt <- melt(scagBinningDf, id=c("t")) %>%
    mutate(nbin = as.numeric(str_sub(variable, start = -2))) %>%
    mutate(idx = str_sub(variable, end = -3))
  save(scagBinningDf, scagMelt, file = "cache/spiralScagBinning.rda")
} else {
  load("cache/spiralScagBinning.rda")
}
ggplot(scagMelt, aes(x=t, y=value)) +
  geom_line(aes(color=factor(nbin))) + 
  facet_wrap(~idx, ncol=1, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE))

@@

if(!file.exists("cache/spiralScagSmoothing.rda")){
  scagSmooth <- getScagSmooth(spiral1000, as.list(t2full))
  scagSmoothMelt <- melt(scagSmooth, id=c("t", "method", "alpha"))
  save(scagSmooth, scagSmoothMelt, file = "cache/spiralScagSmoothing.rda")
} else {
  load("cache/spiralScagSmoothing.rda")
}

>>>>

```{r load}
library(tourr)
library(tidyverse)
library(scagnostics)
library(gridExtra)
library(tictoc) #timer
library(mbgraphic) #Katrins package
library(GGally)
library(geozoo)
library(minerva) #MINE indices
library(kableExtra)
library(forcats)
```

@@

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine (and Spiral) dataset the parameter $x_6$ (and $x_5$) follows a very distinct distribution.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- gather(ds, variable, value)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between a nuisance and a structured projection (solid lines: $n=1000$; dashed: $n=100$). Most of the indexes have peaks at the end of the sequence, indicating that they see the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection."}
fullResMeltPlanned <- fullResPlanned %>%
  mutate(convex = 1-convex) %>% 
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

```{r pipeFirstRun, fig.width=7, fig.height=7, out.width = "0.8\\textwidth", fig.cap="Scouting for the pipe using optimization method 2 on the TIC index. Other PPI values also shown for interest. Only the values for the target plane are shown because the optimization is done off-line. The TIC maximum value for this data is small,  even though it identifies the pipe. The holes, convex, skinny, and to some extent MIC, mirror the performance of TIC for this structure."}
if(!file.exists("cache/pipeFirstRun.rda")){
  pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
  #pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
  #never used?
  #save(pipeTourRes, pipeTourResMelt, file = "cache/pipeFirstRun.rda")
  save(pipeTourRes, file = "cache/pipeFirstRun.rda")
} else {
  load("cache/pipeFirstRun.rda")
}
newBasisPipeTour <- pipeTourRes %>%
  mutate(convex = 1-convex) %>% 
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size) %>%
  filter(t %in% which(attributes(interpolate(pipeTour))$new_basis))
ggplot(newBasisPipeTour, aes(x=t, y=value)) +
  geom_point(mapping=aes(color=(variable=="TIC"))) + 
  geom_line(mapping=aes(color=(variable=="TIC"))) + 
  scale_colour_manual(values = c("black", "red")) +
  facet_wrap(~variable, ncol=1, strip.position = "right") +
  guides(color=FALSE) +
  ylim(c(0,1)) +
  xlab("Number of projections") +
  ylab("PPI value") 
```

```{r refinepipe, results="hide"}
if(!file.exists("cache/refinepipe.rda")){
  iLast <- length(pipeTourFull)
  fProj <- pipeTourFull[[iLast]]
  pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
  pipeTourFull2 <- as.list(interpolate(pipeTour2))
  pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
  #pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
  #save(pipeTour2, pipeTourFull2, pipeTourRes2, pipeTourResMelt2, file = "cache/refinepipe.rda")
  save(pipeTour2, pipeTourFull2, pipeTourRes2, file = "cache/refinepipe.rda")
} else {
  load("cache/refinepipe.rda")
}
```

@@

```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising the splines2d index for the Sine 1000 dataset. Anchor planes are marked by the blue vertical lines.", results="hide"}
set.seed(2018)
if(!file.exists("cache/findsine.rda")){
  sineResc <- as.matrix(sin1000)
  sineTour <- save_history(sineResc,
                          guided_tour(splineIndex()))
  sineTourFull <- as.list(interpolate(sineTour))
  sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
  #sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
  #save(sineResc, sineTour, sineTourFull, sineTourRes, sineTourResMelt, file = "cache/findsine.rda")
  save(sineResc, sineTour, sineTourFull, sineTourRes, file = "cache/findsine.rda")
} else {
  load("cache/findsine.rda")
}
anchorIdx <- as.tibble(which(attributes(interpolate(sineTour))$new_basis))
sineTourResMelt <-sineTourRes %>%
  mutate(convex = 1-convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  select(-cmass) %>%
  gather(variable, value, -t, -name, -size)
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line(mapping=aes(color=(variable=="splines2d"))) + 
  facet_wrap(~variable, ncol=1, scales = "free_y") +
  scale_colour_manual(values = c("black", "red")) +
  facet_wrap(~variable, ncol=1) +
  geom_vline(data=anchorIdx, mapping=aes(xintercept=value), color="blue", alpha=0.3) +
  guides(color=FALSE) +
  ylim(c(0,1))
```

@@

  return(sc)
}
if(!file.exists("cache/spiralScagBinning.rda")){
  scagBinningDf <- getScagComp(spiral1000, t2full)
  scagMelt <- gather(scagBinningDf, variable, value, -t) %>%
    mutate(nbin = as.numeric(str_sub(variable, start = -2))) %>%
    mutate(idx = str_sub(variable, end = -3))
  save(scagBinningDf, scagMelt, file = "cache/spiralScagBinning.rda")
} else {
  load("cache/spiralScagBinning.rda")
}
ggplot(scagMelt, aes(x=t, y=value)) +
  geom_line(aes(color=factor(nbin))) + 
  facet_wrap(~idx, ncol=1, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE))

@@

if(!file.exists("cache/spiralScagSmoothing.rda")){
  scagSmooth <- getScagSmooth(spiral1000, as.list(t2full))
  scagSmoothMelt <- gather(scagSmooth, variable, value, -t, -method, -alpha)
  save(scagSmooth, scagSmoothMelt, file = "cache/spiralScagSmoothing.rda")
} else {
  load("cache/spiralScagSmoothing.rda")
}
